---
title: "General workflow no covariate"
output: html_notebook
---
 CZ529 3a: extend the basic model of CZ325 to 7 managed areas and 7 sub areas of GTMNERR
 Also includes parameters a.k.a responses of Shell Height, Percent Live and Density
 
```{r}
library(tidyverse)
library(magrittr)
library(cmdstanr)
library(here)
library(sf)
library(terra) # is this needed?
library(leaflet) # is this needed?
library(shiny) # is this needed?
library(htmlwidgets)
library(spsurvey) # contains grts
```

```{r}
source(here("./code/SpatialPredictionUtilities.R"))
```

Some shapefiles to partition GTMNERR
```{r}
gtm_managed_area_polygons_Guana_River = st_transform(st_read(here("./data/GTM_Oyster_Monitoring_Region_Polygons"), layer = "Guana_River"), "+proj=longlat +ellps=WGS84 +datum=WGS84")
gtm_managed_area_polygons_Salt_Run =  st_transform(st_read(here("./data/GTM_Oyster_Monitoring_Region_Polygons"), layer = "Salt_Run"), "+proj=longlat +ellps=WGS84 +datum=WGS84")
gtm_managed_area_polygons_Tolomato_River =  st_transform(st_read(here("./data/GTM_Oyster_Monitoring_Region_Polygons"), layer = "Tolomato_River"), "+proj=longlat +ellps=WGS84 +datum=WGS84")
gtm_managed_area_polygons_St_Augustine =  st_transform(st_read(here("./data/GTM_Oyster_Monitoring_Region_Polygons"), layer = "St_Augustine"), "+proj=longlat +ellps=WGS84 +datum=WGS84")
gtm_managed_area_polygons_Pellicer_Flats =  st_transform(st_read(here("./data/GTM_Oyster_Monitoring_Region_Polygons"), layer = "Pellicer_Flats"), "+proj=longlat +ellps=WGS84 +datum=WGS84")
gtm_managed_area_polygons_Butler_Beach =  st_transform(st_read(here("./data/GTM_Oyster_Monitoring_Region_Polygons"), layer = "Butler_Beach"), "+proj=longlat +ellps=WGS84 +datum=WGS84")
gtm_managed_area_polygons_Fort_Matanzas =  st_transform(st_read(here("./data/GTM_Oyster_Monitoring_Region_Polygons"), layer = "Fort_Matanzas"), "+proj=longlat +ellps=WGS84 +datum=WGS84")

```



Input:
1) managed area
2) parameter: one of ShellHeight, PercentLive, Density
3) covariate: one of none, salinity, dissolved_oxygen, depth


"Pine Island Sound Aquatic Preserve" samples has only 3 unique Universal Reef IDs 3036  3039 34843 for ShellHeight
 
"Lemon Bay Aquatic Preserve" has ony 3 unique UniversalReefIDs 3935 29772 29880 for ShellHeight

"Jensen Beach to Jupiter Inlet Aquatic Preserve" has only 1 unique UniversalReefID 5490

"Indian River-Vero Beach to Ft. Pierce Aquatic Preserve" has only 6 unique UniversalReefIDs: 5478  5482  5494  5503  5505 34807
Files required for parameter:  
Uncomment the Managed Area and Parameter you want to model
If the Managed Area is Guana Tolomato Matanzas National Estuarine Research Reserve, select a sub area too
```{r}
# Managed_Area = "Estero Bay Aquatic Preserve" 
 Managed_Area = "Guana Tolomato Matanzas National Estuarine Research Reserve" # 237 points split across regions 
# GTM_sub_area = "Guana_River"
# GTM_sub_area = "Salt_Run"
# GTM_sub_area = "Tolomato_River"
# GTM_sub_area = "St_Augustine"
# GTM_sub_area = "Pellicer_Flats"
 GTM_sub_area = "Butler_Beach"
# GTM_sub_area = "Fort_Matanzas"



# Managed_Area = "Guana River Marsh Aquatic Preserve"
# Managed_Area = "Apalachicola Bay Aquatic Preserve" 
# Managed_Area = "Apalachicola National Estuarine Research Reserve" # includes "Apalachicola Bay Aquatic Preserve" and 17 hrs for shell height
# Managed_Area = "Indian River-Vero Beach to Ft. Pierce Aquatic Preserve" 
## Managed_Area = "Jensen Beach to Jupiter Inlet Aquatic Preserve"
# Managed_Area = "Lemon Bay Aquatic Preserve" 
# Managed_Area = "Pine Island Sound Aquatic Preserve" 



# Parameter = "Shell Height" 
# Parameter = "Percent Live"  
Parameter = "Density" 
Covariate = "None"
scale = ifelse(Parameter == "Density",1/100, 1)

```
Helper function to convert geometry to X,Y
```{r}
ReplaceGeometry = function(SF){
  SF %>%  st_coordinates() %>%  cbind(st_drop_geometry(SF),.)
}
```
# Need to make some labels for the "legacy points"
# first, a litle helper function to add some breaks if the labels get too big
```{r}
add_breaks = function(PL, threshold = 5){
program_ids = PL
program_ids = unlist(str_split(program_ids,","))
if(length(program_ids) <= threshold) return(PL) # else add some breaks
program_chunks <- split(program_ids, ceiling(seq_along(program_ids) / threshold))
program_formatted <- sapply(program_chunks, function(chunk) paste(chunk, collapse = ", "))
program_string <- paste(program_formatted, collapse = "<br>")
}
```

# Managed area geometry
```{r}
SEACAR_GIS_Processing_Layers_ORCP_Managed_Areas_Apr2025 = sf::st_read(dsn = here("./data/SEACAR_GIS_Processing_Layers.gdb"), layer ="ORCP_Managed_Areas_Apr2025") 
SEACAR_GIS_Processing_Layers_ORCP_Managed_Areas_Apr2025 %>%
  filter(LONG_NAME == Managed_Area) %>% 
  sf::st_transform(., crs = st_crs("WGS84")) -> Managed_Area_sf
```

#Sampling locations with all parameters
If "Apalachicola Bay Aquatic Preserve", restrict to 2018 or later. The data contain 200K samples dating back to 1982.
If "Apalachicola National Estuarine Research Reserve", restrict to 2018 or later. The data contain 420K samples dating back to 1982.
```{r}
sf_use_s2(FALSE) #turns off spherical geometry - there are some issues that sf complains about and this fixes them
All_OYSTER_Parameters_2025_Apr_24 = read.table(here("./data/All_OYSTER_Parameters-2025-Apr-24.txt"),   sep = "|", header = TRUE,comment.char = "", quote = "")
All_OYSTER_Parameters_2025_Apr_24 %<>% filter(OriginalLatitude != "NULL")
All_OYSTER_Parameters_2025_Apr_24 %<>% filter(OriginalLongitude != "NULL")
All_OYSTER_Parameters_2025_Apr_24 %<>% mutate( OriginalLatitude = as.numeric(OriginalLatitude))
All_OYSTER_Parameters_2025_Apr_24 %<>% mutate( OriginalLongitude = as.numeric(OriginalLongitude))
All_OYSTER_Parameters_2025_Apr_24_sf = st_as_sf(All_OYSTER_Parameters_2025_Apr_24, coords =  c("OriginalLongitude","OriginalLatitude") , crs = st_crs("WGS84"))
if(Managed_Area == "Apalachicola Bay Aquatic Preserve" || Managed_Area == "Apalachicola National Estuarine Research Reserve"){
All_OYSTER_Parameters_2025_Apr_24_sf %>% filter(ManagedAreaName == Managed_Area) %>% filter(Year >= 2018) %>% filter(ParameterName == Parameter)  -> Managed_Area_Parameter_samples_sf
} else{
  All_OYSTER_Parameters_2025_Apr_24_sf %>% filter(ManagedAreaName == Managed_Area) %>% filter(ParameterName == Parameter)  -> Managed_Area_Parameter_samples_sf
  
}

if( Managed_Area == "Guana Tolomato Matanzas National Estuarine Research Reserve"){
  # choose one of 7 sub areas to intersect with
  if(  GTM_sub_area == "Guana_River"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Guana_River) 
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Guana_River)
     }
  if(  GTM_sub_area == "Salt_Run"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Salt_Run)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Salt_Run)
   }
     
  if(  GTM_sub_area == "Tolomato_River"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Tolomato_River)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Tolomato_River)
  }
  if(  GTM_sub_area == "St_Augustine"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_St_Augustine)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_St_Augustine)
  }
  if(  GTM_sub_area == "Pellicer_Flats"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Pellicer_Flats)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Pellicer_Flats)
  }
  if(  GTM_sub_area == "Butler_Beach"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Butler_Beach)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Butler_Beach)
  }
  
  if(  GTM_sub_area == "Fort_Matanzas"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Fort_Matanzas)  
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Fort_Matanzas)
  }
}


Managed_Area_Parameter_samples_sf %<>% mutate(SampleDate = as.POSIXct(SampleDate))

# Universal Reef ID Lon Lat
# finds the average lon and lat of samples on a reef
cbind(Managed_Area_Parameter_samples_sf, st_coordinates(Managed_Area_Parameter_samples_sf))  %>% dplyr::select( UniversalReefID, SampleDate,X,Y) %>%  group_by(UniversalReefID) %>%  summarize(mean_lon = mean(X), mean_lat = mean(Y) ) %>%
as.data.frame() %>% dplyr::select(UniversalReefID, mean_lon, mean_lat) -> managed_area_parameter_sample_URI_positions 

# add the average positions to the samples.
left_join(Managed_Area_Parameter_samples_sf,managed_area_parameter_sample_URI_positions,by = join_by(UniversalReefID)) %>% 
   group_by(UniversalReefID) %>% mutate(ElapsedMonths = time_length(SampleDate - min(SampleDate), unit = "month"), FirstSampleYear = year(min(SampleDate)), LastSampleYear = year(max(SampleDate)) ) %>%  
   mutate( group = cur_group_id()) -> Managed_Area_Parameter_samples_sf
```

# grid positions for research purposes
```{r}
grid = spsample(Polygon(st_coordinates(Managed_Area_Parameter_samples_sf)[chull(st_coordinates(Managed_Area_Parameter_samples_sf)),]), n = 256, type = "regular")
grid_positions = grid@coords
```

# Oyster positions
```{r}
sf_use_s2(FALSE) #turns off spherical geometry - there are some issues that sf complains about and this fixes them
Oyster_Beds_in_Florida = st_read(here("./data/Oyster_Beds_in_Florida.geojson"))
Oyster_Beds_in_Florida_valid = st_make_valid(Oyster_Beds_in_Florida) # fixes polygons in case there is something wrong with Oyster_Beds_in_Florida polygons
Oyster_Beds_in_Managed_Area = st_intersection(Oyster_Beds_in_Florida_valid, Managed_Area_sf) # these are polygons
Oyster_Beds_in_Managed_Area %>% sf::st_transform(., crs = st_crs("WGS84")) %>% st_point_on_surface() %>% st_point_on_surface() %>% ReplaceGeometry() %>%  dplyr::select(OBJECTID,X,Y) -> managed_area_oyster_objectID_points
```

# Fit spatial random effects model using stan
```{r}

Data_for_stan = list( N =  nrow(Managed_Area_Parameter_samples_sf)
             ,  y = Managed_Area_Parameter_samples_sf$ResultValue*scale
             , G = max(Managed_Area_Parameter_samples_sf$group)
             , group = Managed_Area_Parameter_samples_sf$group
             , positions = (Managed_Area_Parameter_samples_sf %>% as.data.frame() %>% dplyr::select(mean_lon, mean_lat, group) %>% distinct() %>% arrange(group))[,c("mean_lon","mean_lat")] # ordered by group
             )
              


spatial_mean_RE.mod <- cmdstan_model(here("./code/spatial_mean_RE.stan"))

 fit.spatial_mean_RE.Managed_Area_Parameter =  spatial_mean_RE.mod$sample(
    data = Data_for_stan,
    seed = 1111, 
    chains = 4, # 8
    parallel_chains = 4, # 8
    refresh = 1000,
    iter_warmup = 5000, 
    iter_sampling = 5000,
    thin =  1
    #, adapt_delta = 0.99
  )
 
```
# add some measure of prediction

```{r}
data.frame(empirical_mean = (Managed_Area_Parameter_samples_sf %>% group_by(UniversalReefID) %>% summarize( empirical_mean = scale*mean(ResultValue)))$empirical_mean, fitted_mean = fit.spatial_mean_RE.Managed_Area_Parameter$summary("site_mean")$mean) %>% 
  ggplot(aes(x = empirical_mean, y = fitted_mean)) + geom_point() + geom_smooth() + ggtitle(paste(Managed_Area, Parameter)) + geom_abline(intercept = 0, slope = 1)
```





```{r}
Managed_Area_Parameter_samples_sf %>% as.data.frame() %>% dplyr::select(UniversalReefID, FirstSampleYear, LastSampleYear, group) %>% distinct() %>% arrange(group) -> tmp1
Managed_Area_Parameter_samples_sf %>% as.data.frame() %>% dplyr::select(UniversalReefID, ProgramLocationID) %>% distinct() %>% group_by(UniversalReefID) %>%
  summarise(ProgramLocationID = paste(ProgramLocationID, collapse = ", "), .groups = "drop") -> tmp2

tmp3 = inner_join(tmp1, tmp2,by = join_by(UniversalReefID)) %>% arrange(group)%>%group_by(ProgramLocationID) %>% 
  mutate( label = paste("<b>UniversalReefID</b>", UniversalReefID, "<br><b>ProgramLocationID</b>", add_breaks(ProgramLocationID), "<br><b>sampled</b>", FirstSampleYear, "to", LastSampleYear))
shiny_labels = as.vector(tmp3 %>% ungroup() %>% dplyr::select(label))$label
rm(tmp1); rm(tmp2); rm(tmp3)

```

```{r} 
if( Managed_Area == "Guana Tolomato Matanzas National Estuarine Research Reserve"){ # add a sub_area to the filename stem
  filenamestem = paste(gsub(" ","_",Managed_Area), GTM_sub_area, gsub(" ","_",Parameter), sep = "_")
  } else {
    filenamestem = paste(gsub(" ","_",Managed_Area), gsub(" ","_",Parameter), sep = "_")
  }


print(filenamestem)                     
fit.spatial_mean_RE.Managed_Area_Parameter$save_output_files(dir = here("./code/fits/"))  # tell stan to store locally here

# Write out files needed for further analysis and Shiny visualizations

write_rds(Data_for_stan, file = here(paste0("./code/data/",paste(filenamestem,"stan_data.RDS",sep="_")))) # data structure used by Stan
write_rds(Managed_Area_Parameter_samples_sf, file =  here(paste0("./code/data/",paste(filenamestem,"complete_data.RDS",sep="_")))) # complete data
write_rds(fit.spatial_mean_RE.Managed_Area_Parameter, file = here(paste0("./code/fits/",paste(filenamestem,"fit.RDS",sep="_")))) # data structure used by Stan
write_rds(shiny_labels, here(paste0("./code/fits/",paste(filenamestem,"shiny_labels.RDS",sep="_"))))
```
# sd by sample size (per site)
```{r}
sd_by_sample_size = data.frame(sample_size = (Managed_Area_Parameter_samples_sf %>% group_by(UniversalReefID) %>% summarize( n = n()))$n, sd = fit.spatial_mean_RE.Managed_Area_Parameter$summary("site_mean")$sd) %>% arrange(sample_size)

```
```{r}
sd_by_sample_size %>% ggplot(aes(x = sample_size, y = sd/scale)) + geom_jitter(width = 0.1, alpha = 0.2)  + geom_smooth() + ggtitle(" SD by sample size")
```


Prediction on a grid on a convex hull
```{r}
grid.managed_area_parameter.mean_preds = SpatialPrediction(fit = fit.spatial_mean_RE.Managed_Area_Parameter, legacy_positions = Data_for_stan$positions, grid_positions)
grid.managed_area_parameter_positions_sd = data.frame(grid_positions, apply(grid.managed_area_parameter.mean_preds,2,sd))
names(grid.managed_area_parameter_positions_sd) = c("X","Y","sd")

# sd of prediction on oyster positions
oyster.managed_area_parameter.mean_preds = SpatialPrediction(fit = fit.spatial_mean_RE.Managed_Area_Parameter, legacy_positions = Data_for_stan$positions, managed_area_oyster_objectID_points %>% dplyr::select(X,Y))
oyster.managed_area_parameter_positions_sd = data.frame(managed_area_oyster_objectID_points, sd = apply(oyster.managed_area_parameter.mean_preds,2,sd))
names(oyster.managed_area_parameter_positions_sd) = c("OBJECTID","X","Y","sd")

# mean prediction on oyster positions
oyster.managed_area_parameter_positions_mean = data.frame(managed_area_oyster_objectID_points, mean = abs(apply(oyster.managed_area_parameter.mean_preds,2,mean)))
names(oyster.managed_area_parameter_positions_mean) = c("OBJECTID","X","Y","mean")

# coefficient of variation (cv)
oyster.managed_area_parameter_positions_cv = data.frame( OBJECTID = oyster.managed_area_parameter_positions_sd$OBJECTID,
                                                         X = oyster.managed_area_parameter_positions_sd$X,
                                                         Y = oyster.managed_area_parameter_positions_sd$Y,
                                                         cv = oyster.managed_area_parameter_positions_sd$sd/(oyster.managed_area_parameter_positions_mean$mean + 0.1))


# grts sample size is the total sample size including legacy points; there are only so many 
sample_size = min(20 + nrow(Data_for_stan$positions), nrow(managed_area_oyster_objectID_points)) # includes the legacy points.
managed_area_parameter_oyster.positions.sd.sf =st_as_sf(data.frame(managed_area_oyster_objectID_points, sd = apply(oyster.managed_area_parameter.mean_preds,2,sd), var  = apply(oyster.managed_area_parameter.mean_preds,2,var)), coords = c("X","Y"), crs =  st_crs("WGS84"))
grts_new_positions = as.data.frame(st_coordinates(
  grts(managed_area_parameter_oyster.positions.sd.sf,n_base = sample_size, projcrs_check = FALSE, aux_var = "var", legacy_sites = st_as_sf(Data_for_stan$positions, coords = c(1:2), crs =  st_crs("WGS84")) %>% mutate(var = 1))$sites_base)
  )
```


```{r}
system.time(
  (sd_by_number_of_sites = SD_by_number_of_reefs(managed_area_oyster_objectID_points, oyster.managed_area_parameter.mean_preds,legacy_positions = Data_for_stan$positions, fit = fit.spatial_mean_RE.Managed_Area_Parameter))
)

```


```{r}
sd_by_number_of_sites %>% ggplot(aes(x = number_of_sites, y = sd/scale)) + geom_jitter(width = 0.1, alpha = 0.2)  + geom_smooth() + ggtitle("Simulated SD by GRTS sample size with estimation error")
```
```{r}
system.time(
(sd_by_number_of_sites_no_posteriors = SD_by_number_of_reefs_no_posteriors(managed_area_oyster_objectID_points, oyster.managed_area_parameter.mean_preds,legacy_positions = Data_for_stan$positions, fit = fit.spatial_mean_RE.Managed_Area_Parameter))
)
```

```{r}
sd_by_number_of_sites_no_posteriors %>% ggplot(aes(x = number_of_sites, y = sd/scale)) + geom_point()  + geom_smooth() + ggtitle("Simulated SD by GRTS sample size using means only")
```




# Package stuff up for shiny
# "legacy" sample positions
# HTML labels for legacy positions
# Managed Area sf
# Grid positions
# Uncertainty estimates on grid positions
# Oyster bed positions within managed area sf
# Uncertainty estimates on oyster bed positions
# GRTS estimated positions
# sd by sample size

```{r}
shiny_store = list(
Managed_Area = Managed_Area,
Parameter = Parameter,
Covariate = Covariate,  
managed_area_parameter_sample_URI_positions = managed_area_parameter_sample_URI_positions,# "legacy" sample positions
shiny_labels = shiny_labels, # HTML labels for legacy positions
Managed_Area_sf = Managed_Area_sf, # Managed Area sf
grid_positions = grid_positions,
grid.managed_area_parameter_positions_sd = grid.managed_area_parameter_positions_sd,
managed_area_oyster_objectID_points = managed_area_oyster_objectID_points,
Oyster_Beds_in_Managed_Area = Oyster_Beds_in_Managed_Area, 
oyster.managed_area_parameter_positions_sd = oyster.managed_area_parameter_positions_sd,
oyster.managed_area_parameter_positions_mean = oyster.managed_area_parameter_positions_mean,
oyster.managed_area_parameter_positions_cv = oyster.managed_area_parameter_positions_cv,
grts_new_positions = grts_new_positions,
sd_by_sample_size = sd_by_sample_size,
sd_by_number_of_sites = sd_by_number_of_sites,
sd_by_number_of_sites_no_posteriors = sd_by_number_of_sites_no_posteriors,
fit_summary = fit.spatial_mean_RE.Managed_Area_Parameter$summary(c("global_global_mean","length_scale","gp_sigma")) %>% as.data.frame(),
scale = scale)
```

All the data needed for a shiny UI except for the depth map
```{r}
write_rds(shiny_store, file =  here(paste0("./code/data/",paste(filenamestem,"shiny_store.RDS",sep="_"))))
```

