---
title: "R Notebook"
output: html_notebook
---
CZ529 1a; testing covariates on all parameters and covariates; 
9 = 3 covariates x 3 parameters for each managed area
This script evaluates fits from 
a Gaussian Process spatial model
a simple random effects with no spatial component

Covariate:
Salinity
Dissolved oxygen
Bathymetric

Parameter:
Shell Height
Percent Live
Density



```{r}
library(tidyverse)
library(magrittr)
library(sf)
library(terra)
library(here)
library(cmdstanr)
options(mc.cores = parallel::detectCores())
library(spsurvey) # contains grts
library(nngeo)
library(purrr)
source("SpatialPredictionUtilities.R")
```



For a given managed area and parameter (shell height, percent live, density) fit models using all three covariates

```{r}
# Managed_Area = "Estero Bay Aquatic Preserve" # 36 sample sites
# Managed_Area = "Guana River Marsh Aquatic Preserve"
#  Managed_Area = "Apalachicola Bay Aquatic Preserve" 
# Managed_Area = "Apalachicola National Estuarine Research Reserve" # includes "Apalachicola Bay Aquatic Preserve" and 17 hrs for shell height
# Managed_Area = "Indian River-Vero Beach to Ft. Pierce Aquatic Preserve" 
## Managed_Area = "Jensen Beach to Jupiter Inlet Aquatic Preserve"
# Managed_Area = "Lemon Bay Aquatic Preserve" 
# Managed_Area = "Pine Island Sound Aquatic Preserve" 


 Managed_Area = "Guana Tolomato Matanzas National Estuarine Research Reserve" # 237 sample sites; break this into 7 sub areas
# choose one of these
# GTM_sub_area = "Guana_River"
# GTM_sub_area = "Salt_Run"
# GTM_sub_area = "Tolomato_River"
# GTM_sub_area = "St_Augustine"
# GTM_sub_area = "Pellicer_Flats"
# GTM_sub_area = "Butler_Beach"
# GTM_sub_area = "Fort_Matanzas"

#GTM_sub_area = ""
# Managed_Area = "Guana River Marsh Aquatic Preserve"  # 93 samples
# Managed_Area = "Apalachicola Bay Aquatic Preserve"  # 78 samples 
# Managed_Area = "Apalachicola National Estuarine Research Reserve"
# Managed_Area = "Indian River-Vero Beach to Ft. Pierce Aquatic Preserve"
# Managed_Area = "Jensen Beach to Jupiter Inlet Aquatic Preserve" # only one site here so we skip this one
# Managed_Area = "Lemon Bay Aquatic Preserve" 
# Managed_Area = "Pine Island Sound Aquatic Preserve" 


# Parameter = "Shell Height" 
# Parameter = "Percent Live" 
# Parameter = "Density"
scale = 1
if(Parameter == "Density") scale = 1/100 # divide by 100 for density to get the density values within the order of Shell Height and percent live
```



Helper function to replace geometry with simple X and Y
```{r}
ReplaceGeometry = function(SF){
  SF %>%  st_coordinates() %>%  cbind(st_drop_geometry(SF),.)
}
```



```{r}
SEACAR_GIS_Processing_Layers_ORCP_Managed_Areas_Apr2025 = sf::st_read(dsn = here("./data/SEACAR_GIS_Processing_Layers.gdb"), layer ="ORCP_Managed_Areas_Apr2025") 
SEACAR_GIS_Processing_Layers_ORCP_Managed_Areas_Apr2025 %>%
  filter(LONG_NAME == Managed_Area) %>% 
  sf::st_transform(., crs = st_crs("WGS84")) -> Managed_Area_sf

# here are the oyster beds
sf_use_s2(FALSE) #turns off spherical geometry
Oyster_Beds_in_Florida = st_read(here("./data/Oyster_Beds_in_Florida.geojson"))
Oyster_Beds_in_Florida_valid = st_make_valid(Oyster_Beds_in_Florida) # something wrong with Oyster_Beds_in_Florida polygons
Oyster_Beds_in_Managed_Area = st_intersection(Oyster_Beds_in_Florida_valid, Managed_Area_sf) # these are polygons
Oyster_Beds_in_Managed_Area %>% sf::st_transform(., crs = st_crs("WGS84")) %>% st_point_on_surface() %>% dplyr::select(geometry) -> managed_area_oyster_points_sf
Oyster_Beds_in_Managed_Area %>% sf::st_transform(., crs = st_crs("WGS84")) %>% st_point_on_surface() %>% st_coordinates() %>% as.data.frame() %>% dplyr::select(X,Y) -> managed_area_oyster_points 

# here are the samples
All_OYSTER_Parameters_2025_Apr_24 = read.table(here("./data/All_OYSTER_Parameters-2025-Apr-24.txt"),   sep = "|", header = TRUE,comment.char = "", quote = "")
All_OYSTER_Parameters_2025_Apr_24 %<>% filter(OriginalLatitude != "NULL")
All_OYSTER_Parameters_2025_Apr_24 %<>% filter(OriginalLongitude != "NULL")
All_OYSTER_Parameters_2025_Apr_24 %<>% mutate( OriginalLatitude = as.numeric(OriginalLatitude))
All_OYSTER_Parameters_2025_Apr_24 %<>% mutate( OriginalLongitude = as.numeric(OriginalLongitude))
All_OYSTER_Parameters_2025_Apr_24_sf = st_as_sf(All_OYSTER_Parameters_2025_Apr_24, coords =  c("OriginalLongitude","OriginalLatitude") , crs = st_crs("WGS84"))


if(Managed_Area == "Apalachicola Bay Aquatic Preserve" || Managed_Area == "Apalachicola National Estuarine Research Reserve"){
All_OYSTER_Parameters_2025_Apr_24_sf %>% filter(ManagedAreaName == Managed_Area) %>% filter(Year >= 2018) %>% filter(ParameterName == Parameter)  -> Managed_Area_Parameter_samples_sf
} else{
  All_OYSTER_Parameters_2025_Apr_24_sf %>% filter(ManagedAreaName == Managed_Area) %>% filter(ParameterName == Parameter)  -> Managed_Area_Parameter_samples_sf
  
}

if( Managed_Area == "Guana Tolomato Matanzas National Estuarine Research Reserve"){
  # choose one of 7 sub areas to intersect with
  if(  GTM_sub_area == "Guana_River"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Guana_River) 
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Guana_River)
     }
  if(  GTM_sub_area == "Salt_Run"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Salt_Run)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Salt_Run)
   }
     
  if(  GTM_sub_area == "Tolomato_River"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Tolomato_River)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Tolomato_River)
  }
  if(  GTM_sub_area == "St_Augustine"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_St_Augustine)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_St_Augustine)
  }
  if(  GTM_sub_area == "Pellicer_Flats"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Pellicer_Flats)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Pellicer_Flats)
  }
  if(  GTM_sub_area == "Butler_Beach"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Butler_Beach)
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Butler_Beach)
  }
  
  if(  GTM_sub_area == "Fort_Matanzas"){
     Managed_Area_Parameter_samples_sf = st_intersection(Managed_Area_Parameter_samples_sf,gtm_managed_area_polygons_Fort_Matanzas)  
     Managed_Area_sf = st_intersection(Managed_Area_sf,gtm_managed_area_polygons_Fort_Matanzas)
  }
}
#All_OYSTER_Parameters_2025_Apr_24_sf %>% filter(ManagedAreaName == Managed_Area) %>% filter(ParameterName == Parameter)  -> Managed_Area_Parameter_samples_sf
Managed_Area_Parameter_samples_sf %<>% mutate(SampleDate = as.POSIXct(SampleDate))
Managed_Area_Parameter_samples_sf %>% st_coordinates() %>% data.frame() -> Managed_Area_Parameter_samples_points
```


Filter for Salinity 
```{r}
Combined_WQ_WC_NUT_Salinity = read.table(here("./data/Combined_WQ_WC_NUT_Salinity-2025-Mar-06.txt"),   sep = "|", header = TRUE,comment.char = "", quote = "")
Combined_WQ_WC_NUT_Salinity %<>% filter(OriginalLatitude != "NULL")
Combined_WQ_WC_NUT_Salinity %<>% filter(OriginalLongitude != "NULL")
Combined_WQ_WC_NUT_Salinity %<>% mutate( OriginalLatitude = as.numeric(OriginalLatitude))
Combined_WQ_WC_NUT_Salinity %<>% mutate( OriginalLongitude = as.numeric(OriginalLongitude))
Combined_WQ_WC_NUT_Salinity_sf = st_as_sf(Combined_WQ_WC_NUT_Salinity, coords =  c("OriginalLongitude","OriginalLatitude") , crs = st_crs("WGS84"))
Combined_WQ_WC_NUT_Salinity_sf %>% filter(ManagedAreaName == Managed_Area)   -> Managed_Area_Salinity_sf
Managed_Area_Salinity_sf %<>% mutate(SampleDate = as.POSIXct(SampleDate))
Managed_Area_Salinity_sf %>% st_coordinates() %>% data.frame() -> Managed_Area_Salinity_points
```
Filter for Dissolved Oxygen
```{r}
Combined_WQ_WC_NUT_Dissolved_Oxygen = read.table(here("./data/Combined_WQ_WC_NUT_Dissolved_Oxygen-2025-Mar-06.txt"),   sep = "|", header = TRUE,comment.char = "", quote = "")
Combined_WQ_WC_NUT_Dissolved_Oxygen %<>% filter(OriginalLatitude != "NULL")
Combined_WQ_WC_NUT_Dissolved_Oxygen %<>% filter(OriginalLongitude != "NULL")
Combined_WQ_WC_NUT_Dissolved_Oxygen %<>% mutate( OriginalLatitude = as.numeric(OriginalLatitude))
Combined_WQ_WC_NUT_Dissolved_Oxygen %<>% mutate( OriginalLongitude = as.numeric(OriginalLongitude))
Combined_WQ_WC_NUT_Dissolved_Oxygen_sf = st_as_sf(Combined_WQ_WC_NUT_Dissolved_Oxygen, coords =  c("OriginalLongitude","OriginalLatitude") , crs = st_crs("WGS84"))
Combined_WQ_WC_NUT_Dissolved_Oxygen_sf %>% filter(ManagedAreaName == Managed_Area)   -> Managed_Area_Dissolved_Oxygen_sf
Managed_Area_Dissolved_Oxygen_sf %<>% mutate(SampleDate = as.POSIXct(SampleDate))
Managed_Area_Dissolved_Oxygen_sf %>% st_coordinates() %>% data.frame() -> Managed_Area_Dissolved_Oxygen_points
```

Function to extract features of interest for covariates (salinity and dissolved oxygen)
for a given oyster coordinate, find k nearest neighbors
1) min over time and space
2) max over time and space
3) average over time and space
4) min most recent
5) max most recent
6) average most recent

return a set of unique coordinates with the above features
this will work for all oysters and for samples
Oyster samples have a unique UniversalID so we can use that to create an sf
with unique coordinates
```{r}
# make sure sf with Managed Area and Parameter Values are set up with mean coordinates per UniversalReefID
Managed_Area_Parameter_samples_sf %>% ReplaceGeometry() %>% group_by(UniversalReefID) %>% summarize( X = mean(X), Y = mean(Y)) %>% st_as_sf(., coords =  c("X","Y") , crs = st_crs("WGS84")) -> Managed_Area_Parameter_samples_positions_sf

# get unique salinity positions and convert to a sf
Managed_Area_Salinity_sf %>% dplyr::select(geometry) %>% distinct() -> tmp 

# has distinct locations; using 3 nearest neighbors
Samples_to_salinity = nngeo::st_nn(Managed_Area_Parameter_samples_positions_sf, tmp, k = 3, returnDist = TRUE)
positions_in_salinity = purrr::map_df(Samples_to_salinity[[1]], ~as.data.frame(t(.)))
# keep distances to each salinity sample in case we want to filter (not used now)
distances_to_points_in_salinity = purrr::map_df(Samples_to_salinity[[2]], ~as.data.frame(t(.)))


# get salinity coordinates and summarize
# this has all the salinity samples taken at these nn points
# 

# some summary stats; only using avg for now
covariate_stats = NULL
for( ii in 1:nrow(positions_in_salinity)){
tmp1 = inner_join(ReplaceGeometry(Managed_Area_Salinity_sf), ReplaceGeometry(tmp[as.integer(positions_in_salinity[ii,]),]))
covariate_stats = rbind(covariate_stats, 
cbind(tmp1 %>% summarize( min = min(ResultValue), avg = mean(ResultValue), max = max(ResultValue)),
tmp1 %>% filter(SampleDate == max(SampleDate)) %>% summarize( recent_min = min(ResultValue), recent_avg = mean(ResultValue), recent_max = max(ResultValue)))
)
}
Managed_Area_Parameter_samples_positions_covariate_stats_sf = cbind(Managed_Area_Parameter_samples_positions_sf, covariate_stats ) # covariates plus UniversalReefID

# get the stan-ready data frame by left_joining this with the original parameter data
Managed_Area_Parameter_samples_covariates = left_join(ReplaceGeometry(Managed_Area_Parameter_samples_sf), ReplaceGeometry(Managed_Area_Parameter_samples_positions_covariate_stats_sf), by = join_by(UniversalReefID))
Managed_Area_Parameter_samples_covariates %>% group_by(UniversalReefID) %>% mutate(ElapsedMonths = time_length(SampleDate - min(SampleDate), unit = "month"), FirstSampleYear = year(min(SampleDate)), LastSampleYear = year(max(SampleDate)) ) %>%  
   mutate( group = cur_group_id()) %>% dplyr::select(ProgramLocationID, UniversalReefID, FirstSampleYear, LastSampleYear, ResultValue, X.y, Y.y, min, avg, max, recent_min, recent_avg, recent_max, ElapsedMonths, group)  %>% rename("X" = "X.y", "Y" = "Y.y", "covariate" = "avg") -> salinity_data_for_stan_precursor

# for predictions on oyster beds
# get salinity for managed_area_oyster locations in the same way
# Managed_Area_Salinity_sf %>% dplyr::select(geometry) %>% distinct() -> tmp 
# Oysters_to_salinity = st_nn(managed_area_oyster_points_sf, tmp, k = 3, returnDist = TRUE)
# Oyster_positions_in_salinity = purrr::map_df(Oysters_to_salinity[[1]], ~as.data.frame(t(.)))
# Oyster_distances_to_points_in_salinity = purrr::map_df(Oysters_to_salinity[[2]], ~as.data.frame(t(.)))

# # some summary stats
# Oyster_covariate_stats = NULL
# for( ii in 1:nrow(Oyster_positions_in_salinity)){
# tmp1 = inner_join(ReplaceGeometry(Managed_Area_Salinity_sf), ReplaceGeometry(tmp[as.integer(Oyster_positions_in_salinity[ii,]),]))
# Oyster_covariate_stats = rbind(Oyster_covariate_stats, 
# cbind(tmp1 %>% summarize( min = min(ResultValue), avg = mean(ResultValue), max = max(ResultValue)),
# tmp1 %>% filter(SampleDate == max(SampleDate)) %>% summarize( recent_min = min(ResultValue), recent_avg = mean(ResultValue), recent_max = max(ResultValue)))
# )
# }
# 
# # need to add oyster positions geomoetry to these
# Managed_Area_Parameter_oyster_salinity = cbind(managed_area_oyster_points_sf, Oyster_covariate_stats)
```


Fit for salinity
```{r}
tmp_x = as.numeric((salinity_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(covariate, X, Y, group) %>% distinct() %>% arrange(group))[,c("covariate")])
Data_for_stan = list( N =  nrow(salinity_data_for_stan_precursor)
             ,  y = salinity_data_for_stan_precursor$ResultValue*scale
             , x = tmp_x - mean(tmp_x) # centered
             , G = max(salinity_data_for_stan_precursor$group)
             , group = salinity_data_for_stan_precursor$group
             , positions = (salinity_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(X, Y, group) %>% distinct() %>% arrange(group))[,c("X","Y")] # ordered by group
             )

              
spatial_RE.mod <- cmdstan_model(here("./code/spatial_RE.stan")) # allows a covariate on the site mean
 fit.managed_area_parameter_avg_salinity =  spatial_RE.mod$sample(
    data = Data_for_stan,
    seed = 1111, 
    chains = 4, # 8
    parallel_chains = 4, # 8
    refresh = 1000,
    iter_warmup = 5000, 
    iter_sampling = 5000,
    thin =  1
    #, adapt_delta = 0.99
  )
```

```{r}
fit.managed_area_parameter_avg_salinity$summary(c("gp_sigma","length_scale","slope","global_global_mean"))
```
Random effects model with no spatial component

```{r}
tmp_x = as.numeric((salinity_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(covariate, X, Y, group) %>% distinct() %>% arrange(group))[,c("covariate")])
tmp_Data_for_stan = list( N =  nrow(salinity_data_for_stan_precursor)
             ,  y = salinity_data_for_stan_precursor$ResultValue*scale
             , x = tmp_x  - mean(tmp_x) # centered
             , G = max(salinity_data_for_stan_precursor$group)
             , group = salinity_data_for_stan_precursor$group
             )

              
RE.mod <- cmdstan_model(here("./code/RE.stan")) # allows a covariate on the site mean but no spatial information
 fit.managed_area_parameter_avg_salinity_re =  RE.mod$sample(
    data = tmp_Data_for_stan,
    seed = 1111, 
    chains = 4, # 8
    parallel_chains = 4, # 8
    refresh = 1000,
    iter_warmup = 5000, 
    iter_sampling = 5000,
    thin =  1
    #, adapt_delta = 0.99
  )
```

Spatial Model no covariate; compare this to all three covariate models
```{r}
tmp_x = as.numeric((salinity_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(covariate, X, Y, group) %>% distinct() %>% arrange(group))[,c("covariate")]) # not used
tmp_Data_for_stan = list( N =  nrow(salinity_data_for_stan_precursor)
             ,  y = salinity_data_for_stan_precursor$ResultValue*scale
             , G = max(salinity_data_for_stan_precursor$group)
             , group = salinity_data_for_stan_precursor$group
             , positions = (salinity_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(X, Y, group) %>% distinct() %>% arrange(group))[,c("X","Y")] # ordered by group
             )
              


spatial_mean_RE.mod <- cmdstan_model(here("./code/spatial_mean_RE.stan"))

 fit.spatial_mean_RE.Managed_Area_Parameter =  spatial_mean_RE.mod$sample(
    data = tmp_Data_for_stan,
    seed = 1111, 
    chains = 4, # 8
    parallel_chains = 4, # 8
    refresh = 1000,
    iter_warmup = 5000, 
    iter_sampling = 5000,
    thin =  1
    #, adapt_delta = 0.99
  )
 
```


```{r}
fit.lm  = lm(ResultValue~covariate, data = salinity_data_for_stan_precursor)
lm.fitted = data.frame(covariate = salinity_data_for_stan_precursor$covariate, lm.fitted = fit.lm$fitted.values, UniversalReefID = salinity_data_for_stan_precursor$UniversalReefID, group = salinity_data_for_stan_precursor$group)
lm.fitted %<>% group_by(UniversalReefID) %>% summarize( covariate = mean(covariate), lm.pred = mean(lm.fitted)) %>% data.frame()

Empirical_parameter_mean_fitted_covariate = data.frame( UniversalReefID = (salinity_data_for_stan_precursor %>% group_by(UniversalReefID) %>% summarize( covariate_empirical_mean = mean(covariate)))$UniversalReefID
                              , empirical_response_mean = (salinity_data_for_stan_precursor %>% group_by(UniversalReefID) %>% summarize( empirical_response_mean = mean(ResultValue)))$empirical_response_mean
                              , fitted_mean_covariate = fit.managed_area_parameter_avg_salinity$summary("site_mean")$mean/scale
                              , fitted_mean_re = fit.managed_area_parameter_avg_salinity_re$summary("site_mean")$mean/scale
                              , fitted_mean = fit.spatial_mean_RE.Managed_Area_Parameter$summary("site_mean")$mean/scale
                              , fitted_mean_lm  = lm.fitted$lm.pred/scale
                              , covariate = lm.fitted$covariate)
# approximate R^2
cor(Empirical_parameter_mean_fitted_covariate[,2:7])^2
```
Compare spatial with covariate, covariate no spatial, spatial and ordinary linear regression
```{r}

salinity_data_for_stan_precursor %>% ggplot(aes(x = covariate, y = ResultValue)) + geom_jitter(width = 0.1, color = "gray") + 
  geom_jitter(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean), color = "green") + 
  geom_point(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean_covariate), color = "red") + 
  geom_line(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean_re), color = "blue") + 
  geom_point(data = Empirical_parameter_mean_fitted_covariate, aes( x = covariate, y = empirical_response_mean, fill = "orange"), pch=21, size = 3, color = "black") + xlab("Avg site salinity") + ylab(Parameter) + theme_classic() + guides(fill="none")
```
```{r}
fit.managed_area_parameter_avg_salinity$summary(c("global_global_mean","slope", "gp_sigma","length_scale"))
```

```{r}
fit.managed_area_parameter_avg_salinity_re$summary(c("global_global_mean","slope"))
```
```{r}
fit.spatial_mean_RE.Managed_Area_Parameter$summary(c("global_global_mean", "gp_sigma","length_scale"))
```


```{r} 
Covariate = "salinity"
if( Managed_Area == "Guana Tolomato Matanzas National Estuarine Research Reserve"){ # add a sub_area to the filename stem
  filenamestem = paste(gsub(" ","_",Managed_Area), GTM_sub_area, gsub(" ","_",Parameter), Covariate, sep = "_")
  } else {
    filenamestem = paste(gsub(" ","_",Managed_Area), gsub(" ","_",Parameter),Covariate, sep = "_")
  }

print(filenamestem)                     
fit.managed_area_parameter_avg_salinity$save_output_files(dir = here("./code/fits/"))  # tell stan to store locally here
fit.managed_area_parameter_avg_salinity_re$save_output_files(dir = here("./code/fits/")) # tell stan to store locally here
# Write out files needed for further analysis and assessment

write_rds(Data_for_stan, file = here(paste0("./code/data/",paste(filenamestem,"stan_data.RDS",sep="_")))) # data structure used by Stan
write_rds(salinity_data_for_stan_precursor, file = here(paste0("./code/data/",paste(filenamestem,"data_for_stan_precursor.RDS",sep="_"))))
write_rds(Managed_Area_Parameter_samples_sf, file =  here(paste0("./code/data/",paste(filenamestem,"complete_data.RDS",sep="_")))) # complete data
write_rds(fit.managed_area_parameter_avg_salinity, file = here(paste0("./code/fits/",paste(filenamestem,"fit.RDS",sep="_")))) # data structure used by Stan
write_rds(fit.managed_area_parameter_avg_salinity_re, file = here(paste0("./code/fits/",paste(filenamestem,"fit.re.RDS",sep="_")))) # data structure used by Stan
write_rds(Empirical_parameter_mean_fitted_covariate, file = here(paste0("./code/fits/",paste(filenamestem,"Empirical_parameter_mean_fitted_covariate.RDS",sep="_")))) # summary stats
```






Next do average dissolved oxygen
```{r}
# get unique salinity positions and convert to a sf
Managed_Area_Dissolved_Oxygen_sf %>% dplyr::select(geometry) %>% distinct() -> tmp 

# has distinct locations; using 3 nearest neighbors
Samples_to_DO = st_nn(Managed_Area_Parameter_samples_positions_sf, tmp, k = 3, returnDist = TRUE)
positions_in_DO = purrr::map_df(Samples_to_DO[[1]], ~as.data.frame(t(.)))
# keep distances to each salinity sample in case we want to filter (not used now)
distances_to_points_in_DO = purrr::map_df(Samples_to_DO[[2]], ~as.data.frame(t(.)))


# get DO coordinates and summarize
# this has all the DO samples taken at these nn points
# 

# some summary stats; only using avg for now
covariate_stats = NULL
for( ii in 1:nrow(positions_in_DO)){
tmp1 = inner_join(ReplaceGeometry(Managed_Area_Dissolved_Oxygen_sf), ReplaceGeometry(tmp[as.integer(positions_in_DO[ii,]),]))
covariate_stats = rbind(covariate_stats, 
cbind(tmp1 %>% summarize( min = min(ResultValue), avg = mean(ResultValue), max = max(ResultValue)),
tmp1 %>% filter(SampleDate == max(SampleDate)) %>% summarize( recent_min = min(ResultValue), recent_avg = mean(ResultValue), recent_max = max(ResultValue)))
)
}
Managed_Area_Parameter_samples_positions_covariate_stats_sf = cbind(Managed_Area_Parameter_samples_positions_sf, covariate_stats ) # covariates plus UniversalReefID


# get the stan-ready data frame by left_joining this with the original parameter data
Managed_Area_Parameter_samples_covariates = left_join(ReplaceGeometry(Managed_Area_Parameter_samples_sf), ReplaceGeometry(Managed_Area_Parameter_samples_positions_covariate_stats_sf), by = join_by(UniversalReefID))
Managed_Area_Parameter_samples_covariates %>% group_by(UniversalReefID) %>% mutate(ElapsedMonths = time_length(SampleDate - min(SampleDate), unit = "month"), FirstSampleYear = year(min(SampleDate)), LastSampleYear = year(max(SampleDate)) ) %>% mutate( group = cur_group_id()) %>% dplyr::select(ProgramLocationID, UniversalReefID, FirstSampleYear, LastSampleYear, ResultValue, X.y, Y.y, min, avg, max, recent_min, recent_avg, recent_max, ElapsedMonths, group) %>% rename("X" = "X.y", "Y" = "Y.y","covariate"="avg") -> DO_data_for_stan_precursor

# # for predictions on oyster beds
# # get salinity for managed_area_oyster locations in the same way
# Managed_Area_Dissolved_Oxygen_sf %>% dplyr::select(geometry) %>% distinct() -> tmp 
# Oysters_to_DO = st_nn(managed_area_oyster_points_sf, tmp, k = 3, returnDist = TRUE)
# Oyster_positions_in_DO = purrr::map_df(Oysters_to_DO[[1]], ~as.data.frame(t(.)))
# Oyster_distances_to_points_in_DO = purrr::map_df(Oysters_to_DO[[2]], ~as.data.frame(t(.)))
# 
# # some summary stats
# Oyster_covariate_stats = NULL
# for( ii in 1:nrow(Oyster_positions_in_DO)){
# tmp1 = inner_join(ReplaceGeometry(Managed_Area_Dissolved_Oxygen_sf), ReplaceGeometry(tmp[as.integer(Oyster_positions_in_DO[ii,]),]))
# Oyster_covariate_stats = rbind(Oyster_covariate_stats, 
# cbind(tmp1 %>% summarize( min = min(ResultValue), avg = mean(ResultValue), max = max(ResultValue)),
# tmp1 %>% filter(SampleDate == max(SampleDate)) %>% summarize( recent_min = min(ResultValue), recent_avg = mean(ResultValue), recent_max = max(ResultValue)))
# )
# }
# 
# # need to add oyster positions to these
# Managed_Area_Parameter_oyster_DO = cbind(managed_area_oyster_points_sf, Oyster_covariate_stats)
```

Fit for dissolved oxygen
```{r}
tmp_x = as.numeric((DO_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(covariate, X, Y, group) %>% distinct() %>% arrange(group))[,c("covariate")])
Data_for_stan = list( N =  nrow(DO_data_for_stan_precursor)
             ,  y = DO_data_for_stan_precursor$ResultValue*scale
             , x = tmp_x - mean(tmp_x) # center
             , G = max(DO_data_for_stan_precursor$group)
             , group = DO_data_for_stan_precursor$group
             , positions = (DO_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(X, Y, group) %>% distinct() %>% arrange(group))[,c("X","Y")] # ordered by group
             )

 
spatial_RE.mod <- cmdstan_model(here("./code/spatial_RE.stan")) # allows a covariate on the site mean
 fit.managed_area_parameter_avg_DO =  spatial_RE.mod$sample(
    data = Data_for_stan,
    seed = 1111, 
    chains = 4, # 8
    parallel_chains = 4, # 8
    refresh = 1000,
    iter_warmup = 5000, 
    iter_sampling = 5000,
    thin =  1
    #, adapt_delta = 0.99
  )
```

Random effects model with no spatial component

```{r}
tmp_x = as.numeric((DO_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(covariate, X, Y, group) %>% distinct() %>% arrange(group))[,c("covariate")])
tmp_Data_for_stan = list( N =  nrow(DO_data_for_stan_precursor)
             ,  y = DO_data_for_stan_precursor$ResultValue*scale
             , x = tmp_x  - mean(tmp_x) # centered
             , G = max(DO_data_for_stan_precursor$group)
             , group = DO_data_for_stan_precursor$group
             )

              
RE.mod <- cmdstan_model(here("./code/RE.stan")) # allows a covariate on the site mean but no spatial information
 fit.managed_area_parameter_avg_DO_re =  RE.mod$sample(
    data = tmp_Data_for_stan,
    seed = 1111, 
    chains = 4, # 8
    parallel_chains = 4, # 8
    refresh = 1000,
    iter_warmup = 5000, 
    iter_sampling = 5000,
    thin =  1
    #, adapt_delta = 0.99
  )
```


```{r}
fit.lm  = lm(ResultValue~covariate, data = DO_data_for_stan_precursor)
lm.fitted = data.frame(covariate = DO_data_for_stan_precursor$covariate, lm.fitted = fit.lm$fitted.values, UniversalReefID = DO_data_for_stan_precursor$UniversalReefID, group = DO_data_for_stan_precursor$group)
lm.fitted %<>% group_by(UniversalReefID) %>% summarize( covariate = mean(covariate), lm.pred = mean(lm.fitted)) %>% data.frame()

Empirical_parameter_mean_fitted_covariate = data.frame( UniversalReefID = (DO_data_for_stan_precursor %>% group_by(UniversalReefID) %>% summarize( covariate_empirical_mean = mean(covariate)))$UniversalReefID
                              , empirical_response_mean = (DO_data_for_stan_precursor %>% group_by(UniversalReefID) %>% summarize( empirical_response_mean = mean(ResultValue)))$empirical_response_mean
                              , fitted_mean_covariate = fit.managed_area_parameter_avg_DO$summary("site_mean")$mean/scale
                              , fitted_mean_re = fit.managed_area_parameter_avg_DO_re$summary("site_mean")$mean/scale
                              , fitted_mean = fit.spatial_mean_RE.Managed_Area_Parameter$summary("site_mean")$mean/scale
                              , fitted_mean_lm  = lm.fitted$lm.pred/scale
                              , covariate = lm.fitted$covariate)
# approximate R^2
cor(Empirical_parameter_mean_fitted_covariate[,2:7])^2
```
Compare with linear model,  spatial effect
```{r}

DO_data_for_stan_precursor %>% ggplot(aes(x = covariate, y = ResultValue)) + geom_jitter(width = 0.1, color = "gray") + 
  geom_point(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean_covariate), color = "red") +
  geom_point(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean), color = "green") + 
  geom_line(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean_re), color = "blue") + 
  geom_point(data = Empirical_parameter_mean_fitted_covariate, aes( x = covariate, y = empirical_response_mean, fill = "orange"), pch=21, size = 3, color = "black") + xlab("Avg site dissolved oxygen") + ylab(Parameter) + theme_classic() + guides(fill="none")
```


```{r}
fit.managed_area_parameter_avg_DO$summary(c("global_global_mean","slope","gp_sigma", "length_scale"))
```
```{r}
fit.managed_area_parameter_avg_DO_re$summary(c("global_global_mean","slope"))
```
```{r}
fit.spatial_mean_RE.Managed_Area_Parameter$summary(c("global_global_mean", "gp_sigma","length_scale"))
```


```{r}
Covariate = "dissolved_oxygen"
if( Managed_Area == "Guana Tolomato Matanzas National Estuarine Research Reserve"){ # add a sub_area to the filename stem
  filenamestem = paste(gsub(" ","_",Managed_Area), GTM_sub_area, gsub(" ","_",Parameter),Covariate, sep = "_")
  } else {
    filenamestem = paste(gsub(" ","_",Managed_Area), gsub(" ","_",Parameter), Covariate, sep = "_")
  }

print(filenamestem)                     
fit.managed_area_parameter_avg_DO$save_output_files(dir = here("./code/fits/"))  # tell stan to store locally here
fit.managed_area_parameter_avg_DO_re$save_output_files(dir = here("./code/fits/")) # tell stan to store locally here
# Write out files needed for further analysis and assessment

write_rds(Data_for_stan, file = here(paste0("./code/data/",paste(filenamestem,"stan_data.RDS",sep="_")))) # data structure used by Stan
write_rds(DO_data_for_stan_precursor, file = here(paste0("./code/data/",paste(filenamestem,"data_for_stan_precursor",sep="_"))))
write_rds(Managed_Area_Parameter_samples_sf, file =  here(paste0("./code/data/",paste(filenamestem,"complete_data.RDS",sep="_")))) # complete data
write_rds(fit.managed_area_parameter_avg_DO, file = here(paste0("./code/fits/",paste(filenamestem,"fit.RDS",sep="_")))) # data structure used by Stan
write_rds(fit.managed_area_parameter_avg_DO_re, file = here(paste0("./code/fits/",paste(filenamestem,"fit.re.RDS",sep="_")))) # data structure used by Stan
write_rds(Empirical_parameter_mean_fitted_covariate, file = here(paste0("./code/fits/",paste(filenamestem,"Empirical_parameter_mean_fitted_covariate.RDS",sep="_")))) # summary stats
```





Now do Bathymetry

```{r}

Managed_Area_Depth_Map = project(rast(here(paste0("./code/data/",paste(gsub(" ","_",Managed_Area),GTM_sub_area,"full_res_depth_map.tif",sep="_")))),"WGS84")


# get depths at sample points but need to use mean lons lats per UniversalReefID
#Managed_Area_Parameter_samples_sf
Managed_Area_Parameter_samples_sf %>% ReplaceGeometry() %>% group_by(UniversalReefID) %>% summarize( X = mean(X), Y = mean(Y)) %>% st_as_sf(., coords =  c("X","Y") , crs = st_crs("WGS84")) -> Managed_Area_Parameter_samples_positions_sf


#get depths at samples
managed_area_samples_points_depth_sf = cbind(Managed_Area_Parameter_samples_positions_sf,depth = terra::extract( Managed_Area_Depth_Map,Managed_Area_Parameter_samples_positions_sf)[,2])

Managed_Area_Parameter_samples_depth_sf = left_join(ReplaceGeometry(Managed_Area_Parameter_samples_sf), ReplaceGeometry(managed_area_samples_points_depth_sf), by = join_by(UniversalReefID))
Managed_Area_Parameter_samples_depth_sf %>% group_by(UniversalReefID) %>% mutate(ElapsedMonths = time_length(SampleDate - min(SampleDate), unit = "month"), FirstSampleYear = year(min(SampleDate)), LastSampleYear = year(max(SampleDate)) ) %>% mutate( group = cur_group_id()) %>% dplyr::select(ProgramLocationID, UniversalReefID, FirstSampleYear, LastSampleYear, ResultValue, X.y, Y.y, depth, ElapsedMonths, group) %>% rename("X" = "X.y", "Y" = "Y.y", "covariate" = "depth") -> depth_data_for_stan_precursor


```

Fit for depth
```{r}
tmp_x = as.numeric((depth_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(covariate, X, Y, group) %>% distinct() %>% arrange(group))[,c("covariate")])
Data_for_stan = list( N =  nrow(depth_data_for_stan_precursor)
             ,  y = depth_data_for_stan_precursor$ResultValue*scale
             , x = tmp_x - mean(tmp_x)
             , G = max(depth_data_for_stan_precursor$group)
             , group = depth_data_for_stan_precursor$group
             , positions = (depth_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(X, Y, group) %>% distinct() %>% arrange(group))[,c("X","Y")] # ordered by group
             )

 
spatial_RE.mod <- cmdstan_model(here("./code/spatial_RE.stan")) # allows a covariate on the site mean
 
fit.managed_area_parameter_depth =  spatial_RE.mod$sample(
    data = Data_for_stan,
    seed = 1111, 
    chains = 4, # 8
    parallel_chains = 4, # 8
    refresh = 1000,
    iter_warmup = 5000, 
    iter_sampling = 5000,
    thin =  1
    #, adapt_delta = 0.99
  )
```

Random effects model with no spatial component

```{r}
tmp_x = as.numeric((depth_data_for_stan_precursor %>% as.data.frame() %>% dplyr::select(covariate, X, Y, group) %>% distinct() %>% arrange(group))[,c("covariate")])
tmp_Data_for_stan = list( N =  nrow(depth_data_for_stan_precursor)
             ,  y = depth_data_for_stan_precursor$ResultValue*scale
             , x = tmp_x  - mean(tmp_x) # centered
             , G = max(depth_data_for_stan_precursor$group)
             , group = depth_data_for_stan_precursor$group
             )

              
RE.mod <- cmdstan_model(here("./code/RE.stan")) # allows a covariate on the site mean but no spatial information
 fit.managed_area_parameter_depth_re =  RE.mod$sample(
    data = tmp_Data_for_stan,
    seed = 1111, 
    chains = 4, # 8
    parallel_chains = 4, # 8
    refresh = 1000,
    iter_warmup = 5000, 
    iter_sampling = 5000,
    thin =  1
    #, adapt_delta = 0.99
  )
```


```{r}
fit.lm  = lm(ResultValue~covariate, data = depth_data_for_stan_precursor)
lm.fitted = data.frame(covariate = depth_data_for_stan_precursor$covariate, lm.fitted = fit.lm$fitted.values, UniversalReefID = depth_data_for_stan_precursor$UniversalReefID, group = depth_data_for_stan_precursor$group)
lm.fitted %<>% group_by(UniversalReefID) %>% summarize( covariate = mean(covariate), lm.pred = mean(lm.fitted)) %>% data.frame()

Empirical_parameter_mean_fitted_covariate = data.frame( UniversalReefID = (DO_data_for_stan_precursor %>% group_by(UniversalReefID) %>% summarize( covariate_empirical_mean = mean(covariate)))$UniversalReefID
                              , empirical_response_mean = (depth_data_for_stan_precursor %>% group_by(UniversalReefID) %>% summarize( empirical_response_mean = mean(ResultValue)))$empirical_response_mean
                              , fitted_mean_covariate = fit.managed_area_parameter_depth$summary("site_mean")$mean/scale
                              , fitted_mean_re = fit.managed_area_parameter_depth_re$summary("site_mean")$mean/scale
                              , fitted_mean = fit.spatial_mean_RE.Managed_Area_Parameter$summary("site_mean")$mean/scale
                              , fitted_mean_lm  = lm.fitted$lm.pred/scale
                              , covariate = lm.fitted$covariate)
# approximate R^2
cor(Empirical_parameter_mean_fitted_covariate[,2:7])^2
```
Compare with linear model, so spatial effect
```{r}

depth_data_for_stan_precursor %>% ggplot(aes(x = covariate, y = ResultValue)) + geom_jitter(width = 0.01, color = "gray") + 
  geom_jitter(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean), color = "green") + 
  geom_point(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean_covariate), color = "red") +
  geom_line(data = Empirical_parameter_mean_fitted_covariate, aes(x = covariate, y = fitted_mean_re), color = "blue") + 
  geom_point(data = Empirical_parameter_mean_fitted_covariate, aes( x = covariate, y = empirical_response_mean, fill = "orange"), pch=21, size = 3, color = "black") + xlab("Depth") + ylab(Parameter) + theme_classic() + guides(fill="none") +ggtitle(Managed_Area)
```

```{r}
fit.managed_area_parameter_depth$summary(c("global_global_mean","slope","gp_sigma", "length_scale"))
```

```{r}
fit.managed_area_parameter_depth_re$summary(c("global_global_mean","slope"))
```

```{r}
fit.spatial_mean_RE.Managed_Area_Parameter$summary(c("global_global_mean", "gp_sigma","length_scale"))
```


```{r}
Covariate = "depth"
if( Managed_Area == "Guana Tolomato Matanzas National Estuarine Research Reserve"){ # add a sub_area to the filename stem
  filenamestem = paste(gsub(" ","_",Managed_Area), GTM_sub_area, gsub(" ","_",Parameter), Covariate, sep = "_")
  } else {
    filenamestem = paste(gsub(" ","_",Managed_Area), gsub(" ","_",Parameter), Covariate, sep = "_")
  }

print(filenamestem)                     
fit.managed_area_parameter_depth$save_output_files(dir = here("./code/fits/"))  # tell stan to store locally here
fit.managed_area_parameter_depth_re$save_output_files(dir = here("./code/fits/"))  # tell stan to store locally here
#

write_rds(Data_for_stan, file = here(paste0("./code/data/",paste(filenamestem,"stan_data.RDS",sep="_")))) # data structure used by Stan
write_rds(depth_data_for_stan_precursor, file = here(paste0("./code/data/",paste(filenamestem,"data_for_stan_precursor",sep="_"))))
write_rds(Managed_Area_Parameter_samples_sf, file =  here(paste0("./code/data/",paste(filenamestem,"complete_data.RDS",sep="_")))) # complete data
write_rds(fit.managed_area_parameter_depth, file = here(paste0("./code/fits/",paste(filenamestem,"fit.RDS",sep="_")))) # data structure used by Stan
write_rds(fit.managed_area_parameter_depth_re, file = here(paste0("./code/fits/",paste(filenamestem,"fit.re.RDS",sep="_")))) # data structure used by Stan
write_rds(Empirical_parameter_mean_fitted_covariate, file = here(paste0("./code/fits/",paste(filenamestem,"Empirical_parameter_mean_fitted_covariate.RDS",sep="_")))) # summary stats
```


